Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-08-28 papers/article/main.tex
Last pass fixes

%\usepackage{qzapfcha}
%\usepackage{cmbright}
\usepackage{concmath}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgements can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tail} which returns
the rest element of a non-empty list. In \ML{}:
\begin{code}
  let tail = function hd :: tl -> tl | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition tail (l : list A) : option A :=
    match l with
      hd :: tl => Some tl
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : forall p : l <> [] -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. Using \Russell, we would write:
\begin{code}
  Definition tail \{ l : list A | l <> [] \} : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we're defining a partial function and
shows the relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list $tl$ has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $l = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

The remainder of the article is organised as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of translations between
judgements of the two systems in section \ref{sec:Translation}. Then, we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

The subset equivalence idea is to identify elements of subsets if
they are built from the same underlying set. That is, we can identify
elements of $`N$ with elements of any subset $\{ x `: `N `| P \}$.
However, we should not be able to draw conclusions based on the
properties attached to an element of a subset. Otherwise it would be
possible to claim that $0 `: \{ x `: `N `| x \neq 0 \}$ but also that $0
\neq 0$ which would lead us to inconsistency.

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalises this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$`G \vdash P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalises the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalise on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma},
\irule{Sum-Dep}, \irule{Pi-1}, \irule{Pi-2} rules) only when $U, V :
\Set$ or $U, V : \Prop$. The first sum represents dependent pairs,
useful when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is most often used as conjunction
of propositions. The dependent pair with $U : \Set$ and $V : \Prop$ is
the subset type which we distinguish (rule \irule{Subset}). The last
possible pair $U : \Prop, V : \Set$ is forbidden because it corresponds
to a pair where the last component, which is informative, may depend on
a particular proof of the proposition $U$. This is quite contrary to the
mantra ``Computations do not depend on proofs'' which governs our
programming language.

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq 0}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

The reductions rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ it's head normal form.

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

@@

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T'$ and $`G \typea T' \suba T : s$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms. Indeed, the second projection of a
subset element can only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it is forbidden.

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term. We
could also directly use the interpreted types but that complicates the
proof.

@@

\subsection{Properties}
The correctness proof of the translation is very involved, so we will
only sketch it here. A paper version \cite{sozeau:Coq/Russell/report} is
available (in French) and a mechanically checked proof is currently developed.

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and finally that our
interpretation is correct.

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Program Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type $A$.

>>>>

%\usepackage{qzapfcha}
%\usepackage{cmbright}
%\usepackage{concmath}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgements can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term, corresponding to proof obligations (or \PVS{}
  type-checking conditions). A prototype implementation of this process
  has been implemented and integrated with the \Coq{} environment.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent inhabitants of complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tail} which returns
the rest element of a non-empty list. In \ML{}:
\begin{code}
  let tail = function hd :: tl -> tl | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition tail (l : list A) : option (list A) :=
    match l with
      hd :: tl => Some tl
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : forall p : l <> [] -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. Moreover, the user is forced to give a
proof term when calling the function.
We propose a solution to overcome these difficulties, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. Using our new language \Russell, we would write:
\begin{code}
  Definition tail \{ l : list A | l <> [] \} : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we're defining a partial function and
shows the relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list \verb|tl| has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $\texttt{l} = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

The remainder of the article is organised as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
prove decidability of type-checking. Then, we show how it relates to
\CIC{} by means of translations between
judgements of the two systems in section \ref{sec:Translation}. Next we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

In \Coq, the distinction between informative and propositional parts of
a term is formalized by the \Prop/\Set sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalises this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$\vdash_{`G} P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalises the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type-checking condition} $\vdash_{`G} P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalise on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma},
\irule{Sum-Dep}, \irule{Pi-1}, \irule{Pi-2} rules) only when $U, V :
\Set$ or $U, V : \Prop$. The first sum represents dependent pairs,
useful when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is most often used as conjunction
of propositions. The dependent pair with $U : \Set$ and $V : \Prop$ is
the subset type which we distinguish (rule \irule{Subset}). The last
possible pair $U : \Prop, V : \Set$ is forbidden because it corresponds
to a pair where the last component, which is informative, may depend on
a particular proof of the proposition $U$. This is quite contrary to the
mantra ``Computations do not depend on proofs'' which governs our
programming language. If there is no dependence then $U$ and $V$ can be
swapped.

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it won't type-check.
However, in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). These
are not compatible with our subtyping, so we need to forbid the use of
these terms in \Russell{} terms, leaving only the type former (figure
\vref{fig:typing-decl-rules}). With this restriction, when we erase subset types
from \Russell terms, leaving only the supports, we get a valid
\CC term, hence $`_$ is not provable in \Russell.

The reductions rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its head normal form. We use judgemental equality $`G \type T \eqbp U : s$ and the usual untyped
convertibility $\eqbp$ interchangably (we proved $`G \type T \eqbp U : s
"<=>" T \eqbp U `^ `G \type T, U : s$).

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity and symmetry]
  \begin{enumerate}
  \item If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
    \suba V : s$.
  \item If $`G \typea T \suba U : s$ then $`G \typea U \suba T : s$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item By induction on the sum of the depths of the two derivations.
  \item Immediate using a thinning lemma.
  \end{enumerate}
\end{proof}
\end{paragraph}

@@

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then there exists $T'$, $`G \typea t : T'$ and $`G \typea T' \suba T : s$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it is forbidden.

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

@@

\subsection{Properties}
The correctness proof of the translation is very involved, so we will
only sketch it here. A report \cite{sozeau:Coq/Russell/report} is
available (in French) and a mechanically checked proof is under development.

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and the following lemma:
\begin{lemma}[Coercion derivation correctness]
  \label{lemma:coercion-correct}
  If $\subimpl{`G}{c}{T}{U}{s}$ then $\ipG{`G} \typec \lambda x :
  \ip{T}{`G}.c[x] : \ip{T}{`G} "->" \ip{U}{`G}$.
\end{lemma}
 Finally we can show that our interpretation is correct.

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Program Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type
$A$. There is also support for strucural and well-founded \verb|Fixpoint| definitions.
